Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex
Reset the sectioning a bit

The ARG represents the full evolutionary history of SARS-CoV-2 and
is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions in the ARG and the UShER phylogeny (Figure S2).
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
Remarkably, given the differences in inference methods and data processing pipelines,
they are almost identically parsimonious,
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
We also compared the mutational spectra of major VOCs in the ARG
against the mutational spectra computed from the UShER phylogeny (STAR Methods),
finding a close match to previous reports \cite{Bloom2023} (Figure S3).

Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (for more details, see Document S1).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods).
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).

@@

\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}

\section{Major deletions in the ARG}\label{sec:supplement_dels}
% Referenced by the main Results.
Basically, we want to say we did a lot of work to integrate signals of major
indels into the ARG, making it an even more powerful resource overall. Also,
indels are encoded by the Mutation-Annotated Tree format used by UShER
phylogenies (though PanMAT does this).

@@

\section{Analysis of Pango lineage ``origination'' events}

@@

\section{Additional analyses of the Pango Xs}

\subsection{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287
In the ARG, the XS samples appear to have originated from two sequential recombination events.
However, the second event is likely an artifact,
as the XS sample associated with this event failed the minimum net number of supporting loci.
In addition, this sample was sequenced using the AmpliSeq V1 kit,
which was also used to sequence eight other samples descending from the same event.
We excluded this likely artifactual event from further analysis and
consider XS to have emerged from a single recombinant origin.

\subsection{XM}
https://github.com/jeromekelleher/sc2ts-paper/issues/284
Possibly related to deletions.

@@

\subsection*{Analysis of the Jackson et al. recombinants}
Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants
involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
The authors first performed a targeted search for candidate recombinants by
scanning sampled sequences for sequence motifs associated with Alpha, followed
by an analysis to reduce the list of plausible recombinants displaying signs of
onward transmission. Here we compared our sc2ts detection results of these
recombinants with the detection results reported by Jackson et al.

>>>>

The ARG represents the full evolutionary history of SARS-CoV-2 and
is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian 
dataset\cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny\cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions in the ARG and the UShER 
phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
% TODO this is out of date and needs to be rewritten along with supplementary 
% analysis
Remarkably, given the differences in inference methods and data processing pipelines,
they are almost identically parsimonious,
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
We also compared the mutational spectra of major VOCs in the ARG
against the mutational spectra computed from the UShER phylogeny (STAR Methods),
finding a close match to previous reports\cite{Bloom2023}
(Figure~\ref{fig:mutational_spectra}).

Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods).
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).

@@

\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{section}{1}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}

\subsection{Major deletions in the ARG}\label{sec:supplement_dels}
% Referenced by the main Results.
Basically, we want to say we did a lot of work to integrate signals of major
indels into the ARG, making it an even more powerful resource overall. Also,
indels are encoded by the Mutation-Annotated Tree format used by UShER
phylogenies (though PanMAT does this).

@@

\subsection{Analysis of Pango lineage ``origination'' events}

@@

\subsection{Additional analyses of the Pango Xs}

\subsection*{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287
In the ARG, the XS samples appear to have originated from two sequential recombination events.
However, the second event is likely an artifact,
as the XS sample associated with this event failed the minimum net number of supporting loci.
In addition, this sample was sequenced using the AmpliSeq V1 kit,
which was also used to sequence eight other samples descending from the same event.
We excluded this likely artifactual event from further analysis and
consider XS to have emerged from a single recombinant origin.

\subsection*{XM}
https://github.com/jeromekelleher/sc2ts-paper/issues/284
Possibly related to deletions.

@@

\subsection{Analysis of the Jackson et al. recombinants}
Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants
involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
The authors first performed a targeted search for candidate recombinants by
scanning sampled sequences for sequence motifs associated with Alpha, followed
by an analysis to reduce the list of plausible recombinants displaying signs of
onward transmission. Here we compared our sc2ts detection results of these
recombinants with the detection results reported by Jackson et al.
